Covariate analysis of MAYO reprocessed counts (with TMM normalisation)

### 2017-02-13 code/Rmd/Mayo_RNASeq_Reprocessed_TMM_CellTypeFracControl.Rmd
Added markdown to process reprocessed RNASeq from Mayo samples controlled for cell type fraction with TMM normalisation

knit2synapse::knitToFolderEntity(file = "./Mayo_RNASeq_Reprocessed_TMM.Rmd",
                                 parentId = "syn5570325",
                                 entityName = 'UFL_Mayo_ISB Reprocessed RNASeq TMM')
```

@@

```{r synapse.parameters, include=FALSE}
# Synapse parameters
parentId = 'syn5570325';
activityName = 'Covariate analysis';
activityDescription = 'Covariate analysis of UFL_Mayo_ISB RNASeq data';

thisFileName <- 'Mayo_RNASeq_Reprocessed_TMM.Rmd'

@@

METADATA = rbindlist(list(METADATA_TC, METADATA_CE), use.names = T, fill = T) %>%
  as.data.frame
```

@@

rownames(METADATA) = METADATA$ID
METADATA = METADATA[indToRetain,]
```

`r dim(COUNT)[2]` samples from `r length(unique(METADATA$Donor_ID))` subjects were obtained from the Mayo cohorts in AMP-AD reprocessed rnaseq project.

@@

# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.character)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.numeric)
```
Covariate correlation
```{r covariates.correlation, fig.width=8, fig.height=8}
COVARIATES.CORRELATION = CovariateAnalysis::getAssociationStatistics(COVARIATES, PVAL = 0.05)
ggheatmap.show(COVARIATES.CORRELATION$plot, col.width=0.4, row.width=0.1)
```
### Gene Filtering
Preprocess counts matrix and metadata. Determine design matrix for normalisation and differential expression analysis.

@@

# Find significant covariates
adjustCovars = preAdjustedSigCovars$significantCovars
```
Significant covariates to adjust at FDR 0.1 are `r paste(gsub('_','\\\\_',adjustCovars), collapse= ',')`
```{r preAdjustedSigCovars.NULL.ALL, fig.width=25, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```
### Normalisation (iterative)
Since many covariates are correlated, re-normalising COVARIATES with an iterative design matrix.
```{r iterative.adjusted.voom.normalisation, results='asis'}
postAdjustCovars = c('BrainRegion.Diagnosis', 'Source', 'FLOWCELL', 'Gender')
randomCovariates = 'Donor_ID'

# Assign residual covariates
residualSigCovars = preAdjustedSigCovars
significantCovars = residualSigCovars$significantCovars
covariatesEffects = residualSigCovars$Effects.significantCovars
covariatesEffects = covariatesEffects[setdiff(significantCovars, c(postAdjustCovars, randomCovariates))]

postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects)))

loopCount = 0 
while(length(significantCovars)!=0 && loopCount <= 100){
  writeLines(paste('Using following covariates in the model:', 
                   paste(postAdjustCovars, collapse=', '),
                   'as fixed effects'))

  # Post adjusted design matrix
  DM1 = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)
  DM1$design = DM1$design[,linColumnFinder(DM1$design)$indepCols]

  # Estimate voom weights
  VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DM1$design, plot=F)

@@

  # Re-estimate voom weights
  VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DM1$design, plot=F, 
                                       block = COVARIATES$Donor_ID, correlation = correlation$cor)

@@

  # Residual covariates to choose from
  residCovars <- setdiff(c(FactorCovariates,ContCovariates), c(postAdjustCovars, randomCovariates))

  # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                                COVARIATES,
                                                'all adjusted design(voom-normalized)',
                                                isKeyPlot=TRUE)

  # Add postadjusted covariates (if any)
  significantCovars = residualSigCovars$significantCovars
  covariatesEffects = residualSigCovars$Effects.significantCovars
  covariatesEffects = covariatesEffects[setdiff(significantCovars,c(postAdjustCovars, randomCovariates))]

  postAdjustCovars = c(postAdjustCovars,names(which.max(covariatesEffects)))
  loopCount = loopCount + 1
}
modelStr <-paste(paste(gsub('_','\\\\_',postAdjustCovars), collapse=','), 'as fixed effects')

tmp <- paste('Using following covariates in the final model:', modelStr, 'and Donor_ID as random effects')
```
`r tmp`
### Sanity Check
Residualise significant covariates using a linear model and find correlation between PCA of residuals with covariates
```{r residualSigCovars.ALL, fig.width=25, fig.height=12}
residualSigCovars[["PC_res"]][[2]]$plotData
```
Adding "BrainRegion.Diagnosis" back to the residuals
```{r addVarsBack}
varsToAddBackIn <-  grep("BrainRegion.Diagnosis", colnames(FIT$coefficients), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  t(data.matrix(DM1$design[,varsToAddBackIn]) %*% t(FIT$coefficients[,varsToAddBackIn]))
```

@@

VOOM.ADJUSTED.GENE_EXPRESSION$E <- rownameToFirstColumn(VOOM.ADJUSTED.GENE_EXPRESSION$E, 'gene.id') %>%
  left_join(backgroundGenes) %>% 
  left_join(Ensemble2HGNC) %>%
  dplyr::select(gene.id, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
```
### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=10, fig.width=20}
# Fit contrast
contrast = makeContrasts(contrasts=c(
  "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.AD",
  "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.Pathologic_Aging",
  "BrainRegion.DiagnosisCER.Control-BrainRegion.DiagnosisCER.PSP",
  "BrainRegion.DiagnosisCER.Pathologic_Aging-BrainRegion.DiagnosisCER.AD",
  "BrainRegion.DiagnosisCER.Pathologic_Aging-BrainRegion.DiagnosisCER.PSP",
  "BrainRegion.DiagnosisCER.PSP-BrainRegion.DiagnosisCER.AD",
  "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.AD",
  "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.Pathologic_Aging",
  "BrainRegion.DiagnosisTCX.Control-BrainRegion.DiagnosisTCX.PSP",
  "BrainRegion.DiagnosisTCX.Pathologic_Aging-BrainRegion.DiagnosisTCX.AD",
  "BrainRegion.DiagnosisTCX.Pathologic_Aging-BrainRegion.DiagnosisTCX.PSP",
  "BrainRegion.DiagnosisTCX.PSP-BrainRegion.DiagnosisTCX.AD"),
  levels = colnames(FIT$coefficients))
FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

# Get differnetial expression
DE = lapply(1:12, function(i, FIT.CONTR, VOOM.ADJUSTED.GENE_EXPRESSION){
  topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1], confint = T)
}, FIT.CONTR, VOOM.ADJUSTED.GENE_EXPRESSION)
names(DE) = gsub('BrainRegion.Diagnosis','',colnames(contrast))

DE1 = plyr::ldply(DE, .id = 'Comparison') %>%
  dplyr::mutate(Comparison1 = Comparison) %>%
  tidyr::separate(Comparison1, c('from.state','to.state'), sep = '-') %>%
  tidyr::separate(from.state, c('BrainRegion','reference'), sep = '\\.') %>%
  tidyr::separate(to.state, c('BrainRegion','against'), sep = '\\.') %>%
  tidyr::unite(Name, reference, against, sep = '-') %>%
  dplyr::mutate(Comparison = Name, Study = 'MAYO') %>%
  dplyr::rename(gene.id = genes, Region = BrainRegion) %>%
  dplyr::select(-Name) %>%
  left_join(backgroundGenes) %>%
  left_join(Ensemble2HGNC)

p <- ggplot(DE1, aes(x = logFC, y = -log10(adj.P.Val))) + geom_point()
p <- p + geom_hline(yintercept = -log10(0.05), color = 'red')
p <- p + geom_vline(xintercept = log2(1.2), color = 'red')
p <- p + geom_vline(xintercept = -log2(1.2), color = 'red')
p <- p + facet_grid(Region~.+Comparison)
p

tmp = DE1 %>% 
  dplyr::filter(adj.P.Val <= 0.05) %>%
  group_by(Region, Comparison) %>%
  dplyr::summarize(nDiffExpGenes = length(unique(gene.id))) %>%
  spread(Region, nDiffExpGenes)
kable(tmp)
```

### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = 'UFL_Mayo_ISB Reprocessed RNASeq TMM', parentId = parentId)
annotations(CODE) = list(
  dataType = 'mRNA',
  dataSubType = 'geneExp',
  summaryLevel = 'gene',
  assay	 = 'Reprocessed RNAseq',

>>>>

knit2synapse::knitToFolderEntity(file = "./Mayo_RNASeq_Reprocessed_TMM_CellTypeFracControl.Rmd",
                                 parentId = "syn5570325",
                                 entityName = 'UFL_Mayo_ISB Reprocessed RNASeq TMM WithCellTypeFractionControlled')
```

@@

```{r synapse.parameters, include=FALSE}
# Synapse parameters
parentId = 'syn5570325';
activityName = 'Covariate analysis with celltype controlled';
activityDescription = 'Covariate analysis of UFL_Mayo_ISB RNASeq data';

thisFileName <- 'Mayo_RNASeq_Reprocessed_TMM_CellTypeFracControl.Rmd'

@@

METADATA = rbindlist(list(METADATA_TC, METADATA_CE), use.names = T, fill = T) %>%
  as.data.frame

# Get deconvoluted cell type fractions of MAYO samples from synapse 
CELL_TYPE_FRAC = downloadFile('syn8135250') %>%
  dplyr::select(-OPC, -`P-value`, -Correlation, -RMSE)
rownames(CELL_TYPE_FRAC) = CELL_TYPE_FRAC$SampleID
CELL_TYPE_FRAC$SampleID = NULL
CELL_TYPE_FRAC = data.matrix(CELL_TYPE_FRAC)
```

@@

rownames(METADATA) = METADATA$ID
METADATA = METADATA[indToRetain,]

CELL_TYPE_FRAC = CELL_TYPE_FRAC[indToRetain,]
```
`r dim(COUNT)[2]` samples from `r length(unique(METADATA$Donor_ID))` subjects were obtained from the Mayo cohorts in AMP-AD reprocessed rnaseq project.

@@

# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.character)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.numeric)

# Add cell type fractions to the covariates
COVARIATES = cbind(COVARIATES, CELL_TYPE_FRAC[rownames(COVARIATES),])
```
Covariate correlation
```{r covariates.correlation, fig.width=8, fig.height=8}
COVARIATES.CORRELATION = CovariateAnalysis::getAssociationStatistics(COVARIATES, PVAL = 0.05)
ggheatmap.show(COVARIATES.CORRELATION$plot, col.width=0.4, row.width=0.1)
```
cell type fractions
```{r cell.frac, fig.height=6, fig.width=12}
# Cell type fractions
p = CELL_TYPE_FRAC %>%
  rownameToFirstColumn('SampleID') %>%
  left_join(COVARIATES %>%
              rownameToFirstColumn('SampleID') %>%
              dplyr::select(BrainRegion.Diagnosis, SampleID)) %>%
  tidyr::gather(CellType, Fraction, -SampleID, -BrainRegion.Diagnosis) %>%
  ggplot(aes(x = BrainRegion.Diagnosis, y = Fraction, color = CellType))
p = p + geom_boxplot()
p
```

### Gene Filtering
Preprocess counts matrix and metadata. Determine design matrix for normalisation and differential expression analysis.

@@

# Find significant covariates
adjustCovars = preAdjustedSigCovars$significantCovars
```
Significant covariates to adjust at FDR 0.1 are `r paste(gsub('_','\\\\_',adjustCovars), collapse= ',')`
```{r preAdjustedSigCovars.NULL.ALL, fig.width=25, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```
### Normalisation (custom design)
Since many covariates are correlated, re-normalising COVARIATES with a custom design matrix. Here BrainRegion.Diagnosis.Celltype is chose as the primary variable of interest.

Final model:
    Gx = Source + FLOWCELL + Gender + RIN + PMI + BrainRegion.Diagnosis.oligodendrocytes + BrainRegion.Diagnosis.astrocytes + BrainRegion.Diagnosis.microglia + BrainRegion.Diagnosis.neurons + BrainRegion.Diagnosis.endothelial + (1/Donor_ID)

```{r iterative.adjusted.voom.normalisation, results='asis'}
postAdjustCovars = c('BrainRegion.Diagnosis', 'Source', 'FLOWCELL', 'Gender', 'RIN', 'PMI')
randomCovariates = 'Donor_ID'

# Post adjusted design matrix
DESIGN.MATRIX = getDesignMatrix(COVARIATES[,postAdjustCovars,drop=F],Intercept = F)$design

ind.design = grep('BrainRegion.Diagnosis', colnames(DESIGN.MATRIX), value = T)
ind.celltypes = colnames(CELL_TYPE_FRAC)
for(i in ind.celltypes){
  for(j in ind.design){
    tmp = DESIGN.MATRIX[, j, drop = F] * COVARIATES[,i, drop = F]
    colnames(tmp) = paste0(j,'.',i)
    DESIGN.MATRIX = cbind(tmp, DESIGN.MATRIX)
  }
}
DESIGN.MATRIX = DESIGN.MATRIX[, setdiff(colnames(DESIGN.MATRIX), ind.design)]
DESIGN.MATRIX = DESIGN.MATRIX[,linColumnFinder(DESIGN.MATRIX)$indepCols]

writeLines(paste('Using following covariates in the model:', 
                 paste(colnames(DESIGN.MATRIX), collapse=', '),
                 'as fixed effects and Donor_ID as random effect'))

# Estimate voom weights
VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DESIGN.MATRIX, plot=F)

@@

# Re-estimate voom weights
VOOM.ADJUSTED.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DESIGN.MATRIX, plot=F, 
                                     block = COVARIATES$Donor_ID, correlation = correlation$cor)

@@

# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                              COVARIATES,
                                              'all adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)
```
### Sanity Check
Residualise significant covariates using a linear model and find correlation between PCA of residuals with covariates
```{r residualSigCovars.ALL, fig.width=25, fig.height=12}
residualSigCovars[["PC_res"]][[2]]$plotData
```
Adding "BrainRegion.Diagnosis" back to the residuals
```{r addVarsBack}
varsToAddBackIn <-  grep("BrainRegion.Diagnosis", colnames(FIT$coefficients), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  t(data.matrix(DM1$design[,varsToAddBackIn]) %*% t(FIT$coefficients[,varsToAddBackIn]))
```

@@

VOOM.ADJUSTED.GENE_EXPRESSION$E <- rownameToFirstColumn(VOOM.ADJUSTED.GENE_EXPRESSION$E, 'gene.id') %>%
  left_join(backgroundGenes) %>% 
  left_join(Ensemble2HGNC) %>%
  dplyr::select(gene.id, ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
```
### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=10, fig.width=20}
# Fit contrast
contrast = c()
for (region in unique(METADATA$BrainRegion)){
  for (celltype in ind.celltypes){
    contrast = cbind(contrast, 
                     apply(combn(unique(METADATA$Diagnosis),2),2,paste, celltype, sep = '.') %>%
                       apply(2, function(x,y){paste(y,x,sep = '.')}, region))
  }
}
contrast = apply(contrast, 2, function(x,y){paste0(y,x)}, 'BrainRegion.Diagnosis') %>%
  apply(2, paste, collapse = '-')

contrast = makeContrasts(contrasts = contrast, levels = colnames(FIT$coefficients))
FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

# Get differnetial expression
DE = lapply(1:dim(contrast)[2], function(i, FIT.CONTR, VOOM.ADJUSTED.GENE_EXPRESSION){
  topTable(FIT.CONTR, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1], confint = T) %>%
    rownameToFirstColumn('gene.id') %>%
    left_join(backgroundGenes) %>%
    left_join(Ensemble2HGNC) %>%
    dplyr::select(gene.id, ensembl_gene_id, hgnc_symbol, logFC, CI.L, CI.R, AveExpr, t, P.Value, adj.P.Val, B)
}, FIT.CONTR, VOOM.ADJUSTED.GENE_EXPRESSION)
names(DE) = gsub('BrainRegion.Diagnosis','',colnames(contrast))

DE1 = plyr::ldply(DE, .id = 'Comparison') %>%
  tidyr::separate(Comparison, c('c1','c2'), sep = '-') %>%
  tidyr::separate(c1, c('Region', 'cond1', 'celltype'), sep = '\\.') %>%
  tidyr::separate(c2, c('Region2', 'cond2', 'celltype2'), sep = '\\.') %>%
  tidyr::unite(Comparison, cond1, cond2, sep = '_vs_') %>%
  dplyr::select(-celltype2, - Region2) %>%
  dplyr::mutate(Study = 'MAYO',
                Direction = logFC/abs(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')))
DE1$Direction = as.character(DE1$Direction)
DE1$Direction[DE1$logFC < log2(1.2) & DE1$adj.P.Val > 0.05] = 'NONE'
DE1$Direction[DE1$logFC > -log2(1.2) & DE1$adj.P.Val > 0.05] = 'NONE'
DE1$Direction = factor(DE1$Direction, levels = c('DOWN','NONE','UP'))

writeLines('Number of differentially expressed genes at an FDR of 0.05')
DE1 %>%
  dplyr::filter(adj.P.Val <= 0.05) %>%
  dplyr::select(gene.id, Comparison, Region, celltype, Direction) %>%
  group_by(Comparison, Region, celltype, Direction) %>%
  dplyr::summarise(FDR_0_05 = length(unique(gene.id))) %>%
  tidyr::unite('region.celltype.direction', Region, celltype, Direction, sep = '.') %>%
  spread(region.celltype.direction, FDR_0_05) %>%
  kable

writeLines('Number of differentially expressed genes at an FDR of 0.05 and fold change of 1.2')
DE1 %>%
  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
  dplyr::select(gene.id, Comparison, Region, celltype, Direction) %>%
  group_by(Comparison, Region, celltype, Direction) %>%
  dplyr::summarise(FDR_0_05 = length(unique(gene.id))) %>%
  tidyr::unite('region.celltype.direction', Region, celltype, Direction, sep = '.') %>%
  spread(region.celltype.direction, FDR_0_05) %>%
  kable

pl = list()
for (comp in unique(DE1$Comparison)){
  p = ggplot(DE1 %>% filter(Comparison == comp), aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point()
  p = p + scale_color_manual(values = c('green','grey','red'))
  p = p + geom_hline(yintercept = -log10(0.05), color = 'red') 
  p = p + geom_vline(xintercept = log2(1.2), color = 'red') 
  p = p + geom_vline(xintercept = -log2(1.2), color = 'red') 
  p = p + facet_grid(Comparison+Region~.+celltype, scales = 'free_x')
  pl[[comp]] = p
}

pl$AD_vs_Control

pl$PSP_vs_AD

pl$AD_vs_Pathologic_Aging
```

### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = 'UFL_Mayo_ISB Reprocessed RNASeq TMM WithCellTypeFractionControlled', parentId = parentId)
annotations(CODE) = list(
  dataType = 'mRNA',
  dataSubType = 'geneExp',
  summaryLevel = 'gene',
  assay	 = 'Reprocessed RNAseq',
